---
title: "CS-E5710 Bayesian Data Analysis - PROJECT NAME HERE" # TODO: project name here
author: "Egor Eremin & Tam Nguyen"
output: 
  pdf_document:
    toc: yes
    toc_depth: 1
urlcolor: blue
editor_options: 
  markdown: 
    wrap: 72
---

```{r include=FALSE, results='hide'}
# Install/import packages
library(readr)

if (!require(tidybayes)) {
    install.packages("tidybayes")
    library(tidybayes)
}

if (!require(brms)) {
    install.packages("brms")
    library(brms)
}

if(!require(ggplot2)){
    install.packages("ggplot2")
    library(ggplot2)
}

if(!require(cmdstanr)){
    install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
    library(cmdstanr)
}
cmdstan_installed <- function(){
  res <- try(out <- cmdstanr::cmdstan_path(), silent = TRUE)
  !inherits(res, "try-error")
}
if(!cmdstan_installed()){
    install_cmdstan()
}

```

```{r include=FALSE, results='hide'}
# General settings (e.g., seed, package options setting)
seed <- 42

set.seed(seed)
```


```{r}
emission <- read_csv("./data.csv", show_col_types = FALSE)
head(emission)
```

# 3 & 4. Model description & priors: Linear Regression

Using one covariate $RenewableProductionShare$:

$$
\text{CO2EquivEmission} = \alpha + \beta \cdot \text{RenewableProductionShare}
$$

Using one covariate $FossilFuelsPeatProductionShare$:

$$
\text{CO2EquivEmission} = \alpha + \beta \cdot \text{FossilFuelsPeatProductionShare}
$$

Using multiple covariates for different kind of energy production:

$$
\begin{align*}
\text{CO2EquivEmission} 
= \alpha 
&+ \beta_1 \cdot \text{RenewableProductionShare} \\
&+ \beta_2 \cdot \text{FossilFuelsPeatProductionShare} \\
&+ \beta_3 \cdot \text{NuclearEnergyProductionShare} \\
&+ \beta_4 \cdot \text{NetImportsProductionShare} \\
&+ \beta_5 \cdot \text{OthersProductionShare}
\end{align*}

$$


# 5. brms/Stan code & # 6. MCMC inference

How the MCMC inference was run, i.e., what options were used:
- command to run MCMC inference
- a textual explanation of the choice of options

## 5.1. One covariate $RenewableProductionShare$:

```{r}
lin_reg_renewable_formula <- bf(
  CO2EquivEmission ~ 1 + RenewableProductionShare,
  family = "gaussian",
  center = FALSE
)

lin_reg_renewable_priors <- c(
  prior(normal(0, 1000), class = "b", coef = "Intercept"),
  prior(normal(0, 50), class = "b", coef = "RenewableProductionShare"),
  prior(normal(0, 10), class = "sigma")
)

lin_reg_renewable_fit <- brm(
  formula = lin_reg_renewable_formula,
  prior = lin_reg_renewable_priors,
  data = emission,
  iter = 2000,
  warmup = 1000,
  chains = 4,
  seed = seed
)
```

## 5.2. One covariate $FossilFuelsPeatProductionShare$:

```{r}
lin_reg_fossil_fuel_formula <- bf(
  CO2EquivEmission ~ 1 + FossilFuelsPeatProductionShare,
  family = "gaussian",
  center = FALSE
)

lin_reg_fossil_fuel_priors <- c(
  prior(normal(0, 1000), class = "b", coef = "Intercept"),
  prior(normal(0, 50), class = "b", coef = "FossilFuelsPeatProductionShare"),
  prior(normal(0, 10), class = "sigma")
)

lin_reg_fossil_fuel_fit <- brm(
  formula = lin_reg_fossil_fuel_formula,
  prior = lin_reg_fossil_fuel_priors,
  data = emission,
  iter = 2000,
  warmup = 1000,
  chains = 4,
  seed = seed
)
```

## 5.3. Multiple covariates for different kind of energy production:

```{r}
lin_reg_multiple_formula <- bf(
  CO2EquivEmission ~ 1 
                      + RenewableProductionShare
                      + FossilFuelsPeatProductionShare
                      + NuclearEnergyProductionShare
                      + NetImportsProductionShare
                      + OthersProductionShare,
  family = "gaussian",
  center = FALSE
)

lin_reg_multiple_priors <- c(
  prior(normal(0, 1000), class = "b", coef = "Intercept"),  # Broad prior for intercept
  prior(normal(0, 50), class = "b", coef = "RenewableProductionShare"),
  prior(normal(0, 50), class = "b", coef = "FossilFuelsPeatProductionShare"),
  prior(normal(0, 50), class = "b", coef = "NuclearEnergyProductionShare"),
  prior(normal(0, 50), class = "b", coef = "NetImportsProductionShare"),
  prior(normal(0, 50), class = "b", coef = "OthersProductionShare"),
  prior(normal(0, 10), class = "sigma")
)

lin_reg_multiple_fit <- brm(
  formula = lin_reg_multiple_formula,
  prior = lin_reg_multiple_priors,
  data = emission,
  iter = 2000,
  warmup = 1000,
  chains = 4,
  seed = seed
)
```


# 7. Convergence diagnosis

Interpretation of these values:
- R-hat convergence diagnostics
- HMC specific convergence diagnostics (divergences, tree depth)
- ESS diagnostics

If convergence was not good, what did we do to improve?

```{r}
# Convergence, R-hat, ESS all good
summary(lin_reg_renewable_fit)

# I did not test, but I think convergence should be good
summary(lin_reg_fossil_fuel_fit)

# Some divergence, small ESS, this need to be tuned?
summary(lin_reg_multiple_fit)
```


# 8. Posterior predictive checks

What can be interpreted? What was done if the checks indicated misspecifications?

```{r}
pp_check(lin_reg_renewable_fit)

pp_check(lin_reg_fossil_fuel_fit)

pp_check(lin_reg_multiple_fit)
```


# 9. Predictive performance assessment

Absolute error? Relative error? Quantify performance

```{r}
pp_check(lin_reg_renewable_fit, type = "ribbon")

pp_check(lin_reg_fossil_fuel_fit, type = "ribbon")

pp_check(lin_reg_multiple_fit, type = "ribbon")
```

## 9.1. Scatter plot for linear models with one covariate

Renewable model:

```{r}
plot_data <- data.frame(
  RenewableProductionShare = emission$RenewableProductionShare,
  CO2EquivEmission = emission$CO2EquivEmission
)

# Extract posterior predictions for mu and y
mu_predictions <- posterior_epred(lin_reg_fit, re_formula = NA)  # Mean predictions
y_predictions <- posterior_predict(lin_reg_fit)  # Full predictive distribution

# Calculate quantiles for mu
mu_quantiles <- apply(mu_predictions, 2, quantile, probs = c(0.05, 0.5, 0.95))

# Calculate quantiles for y
y_quantiles <- apply(y_predictions, 2, quantile, probs = c(0.05, 0.5, 0.95))

# Create data frames for plotting
mu_plot_data <- data.frame(
  RenewableProductionShare = emission$RenewableProductionShare,
  q5 = mu_quantiles[1, ],
  q50 = mu_quantiles[2, ],
  q95 = mu_quantiles[3, ]
)

y_plot_data <- data.frame(
  RenewableProductionShare = emission$RenewableProductionShare,
  q5 = y_quantiles[1, ],
  q50 = y_quantiles[2, ],
  q95 = y_quantiles[3, ]
)

# Plot
ggplot() +
  geom_point(data = plot_data, aes(x = RenewableProductionShare, y = CO2EquivEmission), color = "blue") +
  # Solid lines for the mean
  geom_line(data = mu_plot_data, aes(x = RenewableProductionShare, y = q50), color = "grey") +
  geom_line(data = y_plot_data, aes(x = RenewableProductionShare, y = q50), color = "red") +
  # Dashed lines for the 95% intervals
  geom_line(data = mu_plot_data, aes(x = RenewableProductionShare, y = q5), color = "grey", linetype = "dashed") +
  geom_line(data = mu_plot_data, aes(x = RenewableProductionShare, y = q95), color = "grey", linetype = "dashed") +
  geom_line(data = y_plot_data, aes(x = RenewableProductionShare, y = q5), color = "red", linetype = "dashed") +
  geom_line(data = y_plot_data, aes(x = RenewableProductionShare, y = q95), color = "red", linetype = "dashed") +
  labs(y = "Life Cycle Emission", x = "Renewable Production Share") +
  theme_minimal()
```

Fossil fuel model:

```{r}
plot_data <- data.frame(
  FossilFuelsPeatProductionShare = emission$FossilFuelsPeatProductionShare,
  CO2EquivEmission = emission$CO2EquivEmission
)

# Extract posterior predictions for mu and y
mu_predictions <- posterior_epred(lin_reg_fit, re_formula = NA)  # Mean predictions
y_predictions <- posterior_predict(lin_reg_fit)  # Full predictive distribution

# Calculate quantiles for mu
mu_quantiles <- apply(mu_predictions, 2, quantile, probs = c(0.05, 0.5, 0.95))

# Calculate quantiles for y
y_quantiles <- apply(y_predictions, 2, quantile, probs = c(0.05, 0.5, 0.95))

# Create data frames for plotting
mu_plot_data <- data.frame(
  FossilFuelsPeatProductionShare = emission$FossilFuelsPeatProductionShare,
  q5 = mu_quantiles[1, ],
  q50 = mu_quantiles[2, ],
  q95 = mu_quantiles[3, ]
)

y_plot_data <- data.frame(
  FossilFuelsPeatProductionShare = emission$FossilFuelsPeatProductionShare,
  q5 = y_quantiles[1, ],
  q50 = y_quantiles[2, ],
  q95 = y_quantiles[3, ]
)

# Plot
ggplot() +
  geom_point(data = plot_data, aes(x = FossilFuelsPeatProductionShare, y = CO2EquivEmission), color = "blue") +
  # Solid lines for the mean
  geom_line(data = mu_plot_data, aes(x = FossilFuelsPeatProductionShare, y = q50), color = "grey") +
  geom_line(data = y_plot_data, aes(x = FossilFuelsPeatProductionShare, y = q50), color = "red") +
  # Dashed lines for the 95% intervals
  geom_line(data = mu_plot_data, aes(x = FossilFuelsPeatProductionShare, y = q5), color = "grey", linetype = "dashed") +
  geom_line(data = mu_plot_data, aes(x = FossilFuelsPeatProductionShare, y = q95), color = "grey", linetype = "dashed") +
  geom_line(data = y_plot_data, aes(x = FossilFuelsPeatProductionShare, y = q5), color = "red", linetype = "dashed") +
  geom_line(data = y_plot_data, aes(x = FossilFuelsPeatProductionShare, y = q95), color = "red", linetype = "dashed") +
  labs(y = "Life Cycle Emission", x = "Fossil Fuels & Peat Production Share") +
  theme_minimal()
```

# 10. Sensitivity analysis

Checking whether the result changes a lot if prior is changed


# 11. Model comparison (e.g. with LOO-CV)

```{r}
loo_compare(
  loo(lin_reg_renewable_fit),
  loo(lin_reg_fossil_fuel_fit),
  loo(lin_reg_multiple_fit)
)
```
