---
title: "CS-E5710 Bayesian Data Analysis - PROJECT NAME HERE" # TODO: project name here
author: "Egor Eremin & Tam Nguyen"
output: 
  pdf_document:
    keep_tex: yes
    toc: yes
    toc_depth: 1
urlcolor: blue
editor_options: 
  markdown: 
    wrap: 72
---

```{r include=FALSE, results='hide'}
# Install/import packages
if (!require(readr)) {
    install.packages("readr")
    library(readr)
}

if (!require(tidybayes)) {
    install.packages("tidybayes")
    library(tidybayes)
}

if (!require(rlang)) {
    install.packages("rlang")
    library(rlang)
}

if (!require(purrr)) {
    install.packages("purrr")
    library(purrr)
}

if (!require(brms)) {
    install.packages("brms")
    library(brms)
}

if(!require(ggplot2)){
    install.packages("ggplot2")
    library(ggplot2)
}

if(!require(ggplot2)){
    install.packages("ggplot2")
    library(ggplot2)
}

if(!require(bayesplot)){
    install.packages("bayesplot")
    library(bayesplot)
}

if(!require(ggpubr)){
    install.packages("ggpubr")
    library(ggpubr)
}

if (!require(loo)) {
  install.packages("loo")
  library(loo)
}

if (!require(priorsense)) {
  install.packages("priorsense")
  library(priorsense)
}

if (!require(reshape2)) {
  install.packages("reshape2")
  library(reshape2)
}

if(!require(cmdstanr)){
    install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
    library(cmdstanr)
}
cmdstan_installed <- function(){
  res <- try(out <- cmdstanr::cmdstan_path(), silent = TRUE)
  !inherits(res, "try-error")
}
if(!cmdstan_installed()){
    install_cmdstan()
}
```

```{r include=FALSE, results='hide'}
# General settings (e.g., seed, package options setting)
seed <- 42

set.seed(seed)
```

```{r include=FALSE, results='hide'}
emission <- read_csv("./data_processing/data/data.csv", show_col_types = FALSE)
```

# 1. Introduction

## 1.1. Motivation

In recent years, the urgency of addressing climate change, particularly global warming, has become increasingly apparent. According to the European Union (EU), achieving carbon neutrality by the mid-21st century is essential to limiting global warming to manageable levels.

To tackle this global challenge, there has been a surge in efforts across various domains, including policy, technological innovation, and international cooperation. For instance, the EU has set a goal to reach carbon neutrality by 2050, while Finland aims to achieve this goal as early as 2035, followed by carbon negativity. Such objectives highlight the necessity of bold and immediate measures to transform energy systems and achieve long-term sustainability.

Renewable energy stands out as one of the most effective tools in this endeavor. A clean energy transition, with an increasing share of renewable energy in energy production and a gradual phasing out of fossil fuels, is central to the strategies of many nations and organizations. 

## 1.2. Problem definition

The CO2-equivalent emission coefficient is a measure used to quantify the carbon emissions associated with energy production. It accounts for the total greenhouse gas emissions, expressed in terms of their CO2-equivalent, per unit of energy generated. This coefficient provides a standardized way to assess and compare the carbon intensity of different energy sources, making it an essential metric for tracking progress toward carbon neutrality.

In this report, we focus on studying Finland's carbon emissions by analyzing its CO2-equivalent emission coefficient for electricity generation over the past two decades. Our aim is to understand the current state of carbon emissions in Finland, examine their historical development, and, if possible, gain insights into the future trajectory. Specifically, we want to explore whether Finland is on track to achieve its 2035 carbon neutrality goal.

Given the critical role of clean energy transition in influencing carbon emissions, we intuitively aim to incorporate data on shares of different energy productions as explanatory variables (covariates), especially production shares of renewables and fossil fuels, hoping that external factors might allow us to better explain the CO2-equivalent emission coefficient and understand the dynamics of carbon emissions in Finland's energy sector.

## 1.3. Modelling idea {#modelling-idea}

Our approach to modeling the CO2-equivalent emission coefficient begins with its nature as a time-series dataset, with yearly observations spanning two decades. Therefore, the first modelling idea are about time-series models, which allow us to analyze the data in terms of trends, seasonality, and cyclical patterns. By using a time-series approach, we aim to let the data speak for itself, revealing how the CO2-equivalent emission coefficient has evolved over time and identifying patterns in its variation.

In parellel to the time-series approach, we also assume that the production shares of different energy sources significantly influence the emission coefficient. Intuitively, a high share of fossil fuel production correlates with higher CO2-equivalent emissions due to the carbon-intensive nature of fossil fuels. Conversely, an increased share of renewable energy production, which emits far less CO2, is expected to reduce the emission coefficient. This logic extends to other energy sources, such as nuclear or imported energy. To quantify these relationships, we frame the problem as a linear regression task, treating the CO2-equivalent emission coefficient as the target variables and the production shares of different energy sources as the covariates. The choice of a linear regression model is straightforward, given that production shares are percentages summing to 100% in each observation, and we assume no inherent non-linear relationship between these shares and emissions.

Finally, recognizing that time-series models often benefit from adding predictive exogenous variables, we plan to integrate insights gained from the regression model into the time-series framework. By incorporating production shares as covariates in the time-series model, we aim to improve its performance and better explain the dynamics between energy production and carbon emissions.

# 2. Data Description and Problem Analysis

## 2.1. Data description

The dataset used in this study is sourced from Statistics Finland, specifically from table [12vp](https://pxdata.stat.fi/PxWeb/pxweb/en/StatFin/StatFin__ehk/statfin_ehk_pxt_12vp.px/) and table [14qt](https://pxdata.stat.fi/PxWeb/pxweb/en/StatFin/StatFin__ehk/statfin_ehk_pxt_14qt.px/). The data was gathered using Python via API calls, followed by preprocessing steps such as merging dataframes and handling missing values.

The dataset spans 23 years, from 2000 to 2022, and contains the following features:

- $RenewableProductionShare$: The share of renewable energy in total energy production, real value in $[0, 100]$ $(\%)$
- $FossilFuelsPeatProductionShare$: The share of fossil fuels and peat in total energy production, real value in $[0, 100]$ $(\%)$
- $NuclearEnergyProductionShare$: The share of nuclear energy in total energy production, real value in $[0, 100]$ $(\%)$
- $NetImportsProductionShare$: The share of imported energy in total energy production, real value in $[0, 100]$ $(\%)$
- $OthersProductionShare$: The share of other energy sources (not classified in the above categories) in total energy production, real value in $[0, 100]$ $(\%)$

The target variable is $CO2EquivEmission$, which measures the CO2-equivalent emissions in real, non-negative values.

No additional normalization or standardization was applied, as the covariates are already expressed as percentages and the target variable is directly interpretable.

## 2.2. Problem Analysis

Despite the public availability of this dataset, we are not aware of any existing studies or analyses specifically examining Finland’s CO2-equivalent emission coefficient in relation to energy production shares. While such studies might exist, we do not actively search for them.
  
# 3. Model Description

In this section, we provide detailed descriptions of the models mentioned in [Section 1.3](#modelling-idea) as follows:


- [Section 3.1](#linreg-def) explains the linear regression model, where production shares of energy sources are used as covariates to estimate their impact on the emission coefficient.
- [Section 3.2](#ts-def) describes the time-series approach, specifically the autoregressive (AR) model.
- [Section 3.3](#hybrid-def) builds on the insights from the linear regression model in [Section 3.2](#linreg-def) to better understand the dynamics between the emission coefficient and production shares. These covariates are then included as exogenous variables in the AR model from [Section 3.1](#ts-def) to improve the performance.

## 3.1. Linear Regression {#linreg-def}


The linear regression model for CO2-equivalent emissions is defined as follows:

$$
y = \alpha + \sum_{i=1}^k \beta_i \cdot x_i + \varepsilon,
$$

where:

- $y$ is the target variable $CO2EquivEmission$,
- $\alpha \in \mathbb{R}$ is the unknown scalar intercept,
- $\beta_i \in \mathbb{R}$ are the unknown scalar slopes for each production share,
- $x_i$ represents the production share of energy source $i$ (e.g., renewable, fossil fuels, nuclear), 
- $\varepsilon \sim \mathcal{N}(0, \sigma^2)$  the noise, and
- $k$ is the total number of energy sources included in the model.

Accordingly, the Bayesian formulation for $CO2EquivEmission$ is

$$
p(y|x_1, ..., x_k,\alpha,\beta_1, ..., \beta_k,\sigma) = p_\mathrm{normal}(y|\alpha + \sum_{i=1}^k \beta_i \cdot x_i, \sigma).
$$

The statistical model further implies

$$
p(y^{\mathrm{pred}}|x_1^{\mathrm{pred}}, ..., x_k^{\mathrm{pred}},\alpha,\beta_1, ..., \beta_k,\sigma) = p_\mathrm{normal}(y|\alpha + \sum_{i=1}^k \beta_i \cdot x_i^{\mathrm{pred}}, \sigma).
$$

as the predictive distribution for a new observation $y^{\mathrm{pred}}$ at a 
given new covariate values $x^{\mathrm{pred}}_1, ..., x^{\mathrm{pred}}_k$.


We will use the following three linear models as explained in [Section 1.3](#modelling-idea):

1. Single Covariate: $RenewableProductionShare$

$$
CO2EquivEmission = \alpha + \beta \cdot RenewableProductionShare
$$


2. Single Covariate: $FossilFuelsPeatProductionShare$

$$
CO2EquivEmission = \alpha + \beta \cdot FossilFuelsPeatProductionShare
$$

3. Multiple Covariates: Full linear model with all types of energy production

$$
\begin{aligned}
CO2EquivEmission
= \alpha 
&+ \beta_1 \cdot RenewableProductionShare \\
&+ \beta_2 \cdot FossilFuelsPeatProductionShare \\
&+ \beta_3 \cdot NuclearEnergyProductionShare \\
&+ \beta_4 \cdot NetImportsProductionShare \\
&+ \beta_5 \cdot OthersProductionShare
\end{aligned}
$$

## 3.2. Time series models {#ts-def}

In addition to the linear models, we will also try using models with an AR component on the CO2 emission coefficient to capture possible "inertia" of CO2 emissions caused by changes in the energy infrastructure, as well as other factors besides the change in the distribution of the energy production methods such as improvements in energy production technology reducing emissions in general.

The plain autoregressive model is given by
$$
y_i = \alpha + \sum_{j=1}^p \phi_j y_{i-j} + \varepsilon 
$$
where:
- $y_i$ is the $i$th observation of the target variable $CO2EquivEmission$,
- $\alpha \in \mathbb{R}$ is the unknown scalar intercept,
- $\phi_j \in \mathbb{R}$ are the unknown regression coefficients for each lag,
- $\varepsilon \sim \mathcal{N}(0, \sigma^2)$ the noise, and
- $p$ is the order of the autoregressive model.

The mean of a purely autoregressive model is given by:
$$
\text{CO2EquivEmission} = \alpha + AR(1)
$$

## 3.3. Linear models with autoregressive component

The Bayesian formulation of the rest of the models is practically the same as that for the linear models, having added AR(1) component to the mean of the model distribution. 

We are going to add only an AR component of order 1, i.e. we use the previous year's observation to predict the next one. The models' means are then given by:

Using one covariate $RenewableProductionShare$ and an AR(1):

$$
\text{CO2EquivEmission} = \alpha + \beta \cdot \text{RenewableProductionShare} + AR(1)
$$

Using one covariate $FossilFuelsPeatProductionShare$ and an AR(1):

$$
\text{CO2EquivEmission} = \alpha + \beta \cdot \text{FossilFuelsPeatProductionShare} + AR(1)
$$

We use the same prior choices as for the linear models. The AR regression coefficient has the following weakly informative prior:
$$
\phi \sim \mathcal{N}(0, 1)
$$


# 4. Prior justification

## 4.1. Linear Regression {#linreg-prior}

The prior for the regression coefficients $\beta_i$ is specified as:

$$
\beta_1, \beta_2, \dots, \beta_k \sim \mathcal{N}(0, 10).
$$

This reflects a weakly informative prior with a standard deviation of 10, allowing 
for a wide range of plausible values for $\beta_i$. By using large variance, 
the prior expresses minimal prior knowledge, given that we do not have much
information about plausible values for $\beta_i$, and we would like the data
to speak for itself.


The prior for the intercept $\alpha$ is specified as:

$$
\alpha \sim \mathcal{N}(0, 500).
$$

Since $y$ has a large range, we assign a prior with a large standard deviation 
to reflect the broader uncertainty about the baseline value of $y$ when all 
covariates are zero. The large variance ensures that the prior is weakly informative 
and does not overly constrain the model.


The prior for $\sigma$ is given by:

$$
\sigma \sim \mathcal{N}^+(0, 10),
$$

where $\mathcal{N}^+$ indicates the positive half-normal distribution. This ensures 
that $\sigma$ is strictly positive and reflects weak prior knowledge about the 
variability in the observations.

## 4.2. Autoregressive models {#ts-prior}

For the models with the autoregressive component, we use the same prior choices as for the linear models. The AR regression coefficient has the following weakly informative prior:
$$
\phi \sim \mathcal{N}(0, 1)
$$

# 5. brms/rstanarm/Stan code

We use `brms` to fit our models. In addition to the `brms` formulas, we also
define the priors for each model according to our prior justifications above.

## 5.1. Linear Regression {#linreg-code}

### 5.1.1. Single covariate $RenewableProductionShare$:

```{r results='hide'}
lin_reg_renewable_formula <- bf(
  CO2EquivEmission ~ 1 + RenewableProductionShare,
  family = "gaussian",
  center = FALSE
)

lin_reg_renewable_priors <- c(
  prior(normal(0, 500), class = "b", coef = "Intercept"),
  prior(normal(0, 10), class = "b", coef = "RenewableProductionShare"),
  prior(normal(0, 10), class = "sigma")
)
```

### 5.1.2. One covariate $FossilFuelsPeatProductionShare$:

```{r results='hide'}
lin_reg_fossil_fuel_formula <- bf(
  CO2EquivEmission ~ 1 + FossilFuelsPeatProductionShare,
  family = "gaussian",
  center = FALSE
)

lin_reg_fossil_fuel_priors <- c(
  prior(normal(0, 500), class = "b", coef = "Intercept"),
  prior(normal(0, 10), class = "b", coef = "FossilFuelsPeatProductionShare"),
  prior(normal(0, 10), class = "sigma")
)
```

### 5.1.3. Multiple covariates for different kind of energy production:

```{r results='hide'}
lin_reg_multiple_formula <- bf(
  CO2EquivEmission ~ 1 
                      + RenewableProductionShare
                      + FossilFuelsPeatProductionShare
                      + NuclearEnergyProductionShare
                      + NetImportsProductionShare
                      + OthersProductionShare,
  family = "gaussian",
  center = FALSE
)

lin_reg_multiple_priors <- c(
  prior(normal(0, 500), class = "b", coef = "Intercept"),  # Broad prior for intercept
  prior(normal(0, 10), class = "b", coef = "RenewableProductionShare"),
  prior(normal(0, 10), class = "b", coef = "FossilFuelsPeatProductionShare"),
  prior(normal(0, 10), class = "b", coef = "NuclearEnergyProductionShare"),
  prior(normal(0, 10), class = "b", coef = "NetImportsProductionShare"),
  prior(normal(0, 10), class = "b", coef = "OthersProductionShare"),
  prior(normal(0, 10), class = "sigma")
)
```

## 5.2. Autoregressive models {#ts-code}

## 5.1 AR(1):
```{r results='hide'}
ar_formula <- bf(
  CO2EquivEmission ~ ar(time = Year, p=1), family="Gaussian"
)
ar_priors <- c(
  prior(normal(0, 1), class = "ar"),
  prior(normal(0, 500), class="Intercept"),
  prior(normal(0, 10), class = "sigma")
)
```

## 5.2. One covariate $RenewableProductionShare$ and AR(1):

```{r results='hide'}
ar_renewable_formula <- bf(
  CO2EquivEmission ~ RenewableProductionShare + ar(time = Year, p=1),
  family = "gaussian",
  center = FALSE
)

ar_renewable_priors <- c(
  prior(normal(0, 500), class = "b", coef = "Intercept"),
  prior(normal(0, 10), class = "b", coef = "RenewableProductionShare"),
  prior(normal(0, 10), class = "sigma"),
  prior(normal(0, 1), class = "ar")
  )
```

## 5.3. One covariate $FossilFuelsPeatProductionShare$ and AR(1): 

```{r results='hide'}
ar_fossil_fuel_formula <- bf(
  CO2EquivEmission ~ FossilFuelsPeatProductionShare + ar(time = Year, p=1),
  family = "gaussian",
  center = FALSE
)

ar_fossil_fuel_priors <- c(
  prior(normal(0, 500), class = "b", coef = "Intercept"),
  prior(normal(0, 10), class = "b", coef = "FossilFuelsPeatProductionShare"),
  prior(normal(0, 10), class = "sigma"),
  prior(normal(0, 1), class = "ar")
)
```


# 6. MCMC inference

The models are run with 4 chains, 2000 iterations, and 1000 warm-ups.   

## 6.1. Linear Regression {#linreg-inference}

```{r results='hide'}
lin_reg_renewable_fit <- brm(
  formula = lin_reg_renewable_formula,
  prior = lin_reg_renewable_priors,
  data = emission,
  iter = 2000,
  warmup = 1000,
  chains = 4,
  seed = seed
)

lin_reg_fossil_fuel_fit <- brm(
  formula = lin_reg_fossil_fuel_formula,
  prior = lin_reg_fossil_fuel_priors,
  data = emission,
  iter = 2000,
  warmup = 1000,
  chains = 4,
  seed = seed
)

lin_reg_multiple_fit <- brm(
  formula = lin_reg_multiple_formula,
  prior = lin_reg_multiple_priors,
  data = emission,
  iter = 2000,
  warmup = 1000,
  chains = 4,
  seed = seed
)
```

## 6.2. Autoregressive models {#ts-inference}
```{r results='hide'}
ar_fit <- brm(
  formula = ar_formula, 
  prior = ar_priors, 
  data=emission, 
  iter = 2000,
  warmup = 1000,
  chains = 4,
  seed = seed
  )

ar_renewable_fit <- brm(
  formula = ar_renewable_formula,
  prior = ar_renewable_priors,
  data = emission,
  iter = 2000,
  warmup = 1000,
  chains = 4,
  seed = seed
)

ar_fossil_fuel_fit <- brm(
  formula = ar_fossil_fuel_formula,
  prior = ar_fossil_fuel_priors,
  data = emission,
  iter = 2000,
  warmup = 1000,
  chains = 4,
  seed = seed,
  control=list(adapt_delta = 0.99)
)
```


# 7. Convergence diagnosis

According to the MCMC inference summary below, there is no sign of divergence in most of the models, as we had no divergent transitions, and the $\hat{R}$ values for all of our models are 1. Even though the effective samples sizes differ from the number of samples significantly, they remain sufficiently large, indicating that we still have a good amount of independent samples to use for analysis.

```{r echo=FALSE}
summary(ar_fit)
summary(ar_renewable_fit)
summary(ar_fossil_fuel_fit)
summary(lin_reg_renewable_fit)
summary(lin_reg_fossil_fuel_fit)
summary(lin_reg_multiple_fit) # Divergence problem
```

The only exception is the full linear regression model that uses all available
covariates. For this model, the MCMC summary indicated divergent transitions. 
Using the `pairs()` plot, we diagnosed that the divergence was caused by high 
multicollinearity in the predictors. Specifically, $RenewableProductionShare$ and $FossilFuelsPeatProductionShare$ were found to be highly negatively correlated. 
To resolve this, we removed $RenewableProductionShare$ from the list of covariates.

```{r, echo=FALSE}
# Create a pairs plot for the problematic covariates
pairs(emission[, c("RenewableProductionShare", "FossilFuelsPeatProductionShare")])
```

As a result, we use a linear regression with these covariates:

$$
\begin{aligned}
CO2EquivEmission
= \alpha 
&+ \beta_1 \cdot FossilFuelsPeatProductionShare \\
&+ \beta_2 \cdot NuclearEnergyProductionShare \\
&+ \beta_3 \cdot NetImportsProductionShare \\
&+ \beta_4 \cdot OthersProductionShare
\end{aligned}
$$

Additionally, the estimated values for the intercept and slopes were in a much
smaller range than the prior distributions. Therefore, we updated the prior 
variances to be more informative, yet still weakly informative.

Correspondingly, the updated `brms` model is:

```{r results='hide'}
lin_reg_multiple_fix_formula <- bf(
  CO2EquivEmission ~ 1 
                      + FossilFuelsPeatProductionShare
                      + NuclearEnergyProductionShare
                      + NetImportsProductionShare
                      + OthersProductionShare,
  family = "gaussian",
  center = FALSE
)

lin_reg_multiple_fix_priors <- c(
  prior(normal(0, 250), class = "b", coef = "Intercept"),  # Broad prior for intercept
  prior(normal(0, 5), class = "b", coef = "FossilFuelsPeatProductionShare"),
  prior(normal(0, 5), class = "b", coef = "NuclearEnergyProductionShare"),
  prior(normal(0, 5), class = "b", coef = "NetImportsProductionShare"),
  prior(normal(0, 5), class = "b", coef = "OthersProductionShare"),
  prior(normal(0, 10), class = "sigma")
)
```

```{r, echo=FALSE, results='hide'}
lin_reg_multiple_fix_fit <- brm(
  formula = lin_reg_multiple_formula,
  prior = lin_reg_multiple_priors,
  data = emission,
  iter = 2000,
  warmup = 1000,
  chains = 4,
  seed = seed
)
```

After removing problematic covariate $RenewableProductionShare$ and updating
the priors, we can see that the linear regression model with multiple covariates
does not have divergence problem anymore.

```{r, echo=FALSE}
summary(lin_reg_multiple_fix_fit)
```

# 8. Posterior Predictive Checks

Despite the MCMC inference converging, the density overlay revealed several 
interesting insights during the posterior predictive checks.

First, we observe that the linear model with renewable production share does not 
fit the data well. The density of $y_\text{replicate}$ does not resemble the 
density of the true data $y$, indicating that the model struggles to capture 
the underlying data distribution.

Second, the linear model with fossil fuel production share shows a better fit. 
However, the posterior predictive density of $y_\text{replicate}$ exhibits a 
bimodal distribution, whereas the true data is unimodal. This discrepancy suggests 
that the linear model might fail to capture certain patterns in the data.

Lastly, the linear model with multiple covariates behaves similarly to the linear 
model with only fossil fuels. It also exhibits a bimodal distribution for \(y_\text{replicate}\). This implies that fossil fuels might be the dominant 
predictor for the emission coefficient, with its effect outweighing other 
covariates. Additionally, this suggests that there might be non-linear relationships 
in the data that are not captured by the linear models.

```{r, echo=FALSE}
ggarrange(
  pp_check(lin_reg_renewable_fit) + labs(title="LinReg Renewable"), 
  pp_check(lin_reg_fossil_fuel_fit)+ labs(title="LinReg FossilFuels"), 
  pp_check(lin_reg_multiple_fit)+ labs(title="LinReg Multiple"), 
  common.legend=TRUE, 
  ncol=3)
```

We can see that the plain AR model, expectedly, performed the worst: Some draws indicate bimodal distribtuions, while other draws have too light tails and overestimate density for the middle values. 

AR model with a renewable fuel share covariate performs significantly better as its posterior matches that of data more closely. We still have a problem with tails, where the left side of the posterior is too heavy-tailed and the right side still undersetimates thickness for some draws. We also still observe overestimation of the middle values density in the posterior. 

AR model with the fossil fuel covariate shows the best fit of the posterior to data. The model still produces a bimodal distribution sometimes, as well as still underestimates the right tail values. 

```{r, echo=FALSE}
ggarrange(
 pp_check(ar_fit) + labs(title="AR(1)"), 
 pp_check(ar_renewable_fit)+ labs(title="AR(1) + Renewable"), 
 pp_check(ar_fossil_fuel_fit)+ labs(title="AR(1) + FossilFuels"), 
 common.legend=TRUE, 
 ncol=3)
```


# 9. Predictive performance assessment

The observations in Section 8 above are further confirmed with the time series predictive plots. Below you can find the plots for the models' predicitons. The light blue line is the mean, dark-blue shaded area shows the 50% CI and the light-blue area shows the 90% CI.

First, we can see that the linear model with fossil fuels fits very well to the 
original time series. The predictions fitted values $y_{\text{replicate}}$ closely 
align with the observed data $y$ across most data points, and the credible intervals 
are relatively narrow, reflecting the strong predictive power of fossil fuel 
production share as a covariate.

In contrast, the linear model with renewable production share does not fit as 
well. The credible intervals are wider, and the predicted values $y_{\text{replicate}}$ 
deviate more significantly from the observed data $y$. This supports the conclusion 
that renewable production share alone does not adequately capture the variability in CO2 emissions.

```{r, echo=FALSE}
ggarrange(
  pp_check(lin_reg_renewable_fit, type = "ribbon") + labs(title="LinReg Renewable"), 
  pp_check(lin_reg_fossil_fuel_fit, type = "ribbon")+ labs(title="LinReg FossilFuels"), 
  pp_check(lin_reg_multiple_fit, type = "ribbon")+ labs(title="LinReg Multiple"), 
  common.legend=TRUE, 
  ncol=3)
```

Below you can find the plots for the models' predicitons. The light blue line is the mean, dark-blue shaded area shows the 50% CI and the light-blue area shows the 90% CI. As you can see, the ribbon plots tell us the same story as the posterior density plots above in terms of the models' performance. 

You can also see that the AR(1) model almost reduces to a lag model, with an additional downward trend. The model with the renewable energy production share covariate follows the data better but has still room for improvement. The last model with the fossil fuel covariate fits data the best, which confirms the intuition that the fossil fuel share is the best predictor for the CO2 emission coefficient. 

```{r, echo=FALSE}
ggarrange(
 pp_check(ar_fit, type = "ribbon") + labs(title="AR(1)"), 
 pp_check(ar_renewable_fit, type = "ribbon")+ labs(title="AR(1) + Renewable"), 
 pp_check(ar_fossil_fuel_fit, type = "ribbon")+ labs(title="AR(1) + FossilFuels"), 
 common.legend=TRUE, 
 ncol=3)
```

# 10. Sensitivity analysis

We create a function `prior_check()` that allows us to compare the posterior
distributions of all parameters given different priors.

```{r, results='hide'}
prior_check <- function(priors, model) {
  draws <- priors %>%
  map(~ as.matrix(update(model, prior=.x, seed=seed, refresh=0)))
  
  draws_to_plot <- function(draws) {
    params <- head(colnames(draws[[1]]), -2)
    data <- list()
    for (p in params) {
        for (name in names(draws)) {
          t <- draws[[name]][,p]
          for (s in 1:length(t)) {
            data[[p]][[name]][[s]] = t[[s]]
          }
        }
    }
    data
  }
  dt <- draws_to_plot(draws)
  
  plot_for_loop <- function(df, x_var) {
    df.long <- melt(df)
      ggplot(df.long, aes(x= value, fill=L1)) +
      geom_density(alpha=0.7) +
      labs(x = x_var) +
      theme_classic(base_size = 12)
  }
  
  plot_list <- list()
  params <- head(colnames(draws[[1]]), -2)
  for (p in params) {
    plot_list[[length(plot_list)+1]] <- plot_for_loop(dt[[p]], p)
  }
  
  # Combine all plots
  ggarrange(plotlist=plot_list, common.legend=TRUE)
}
```

## 10.1. Linear Regression {#linreg-sensitivity}

### 10.1.1. Single covariate $RenewableProductionShare$:

```{r}
priors_lin_reg_renewable <- list(
  i500_renewable10_sigma10 = c(
    prior(normal(0, 500), class = "b", coef = "Intercept"),
    prior(normal(0, 10), class = "b", coef = "RenewableProductionShare"),
    prior(normal(0, 10), class = "sigma")
  ),
  i50_renewable1_sigma1 = c(
    prior(normal(0, 50), class = "b", coef = "Intercept"),
    prior(normal(0, 1), class = "b", coef = "RenewableProductionShare"),
    prior(normal(0, 1), class = "sigma")
  ),
  i150_100_renewable5_sigma5 = c(
    prior(normal(150, 100), class = "b", coef = "Intercept"),
    prior(normal(0, 5), class = "b", coef = "RenewableProductionShare"),
    prior(normal(0, 5), class = "sigma")
  )
)

plot <- prior_check(priors_lin_reg_renewable, lin_reg_renewable_fit)
```

From the density plots of the posterior distributions under different priors, we 
observe the following:
- All parameters are quite sensitive to different priors, especially for more informative priors.
- The model's posterior estimate for the intercept is moderately sensitive to the prior, particularly for priors with stronger assumption for mean.
- The posterior for `RenewableProductionShare` is sensitive to the prior variance, indicating strong data influence.
- The posterior for `sigma` is highly sensitive to the prior variance, indicating weak data influence and high reliance on prior information.

### 10.1.2. One covariate $FossilFuelsPeatProductionShare$:

```{r}
priors_lin_reg_fossil_fuel <- list(
  i500_fossil10_sigma10 = c(
    prior(normal(0, 500), class = "b", coef = "Intercept"),
    prior(normal(0, 10), class = "b", coef = "FossilFuelsPeatProductionShare"),
    prior(normal(0, 10), class = "sigma")
  ),
  i50_fossil1_sigma1 = c(
    prior(normal(0, 50), class = "b", coef = "Intercept"),
    prior(normal(0, 1), class = "b", coef = "FossilFuelsPeatProductionShare"),
    prior(normal(0, 1), class = "sigma")
  ),
  i150_100_fossil5_sigma5 = c(
    prior(normal(150, 100), class = "b", coef = "Intercept"),
    prior(normal(0, 5), class = "b", coef = "FossilFuelsPeatProductionShare"),
    prior(normal(0, 5), class = "sigma")
  )
)

prior_check(priors_lin_reg_fossil_fuel, lin_reg_fossil_fuel_fit)
```

From the density plots of the posterior distributions under different priors, we observe the following:
- All parameters are still somewhat sensitive to very small prior variances (i.e., much more informative priors), indicating that the choice of prior can still influence the posterior when priors are overly restrictive.
- The intercept and slope for FossilFuelsPeatProductionShare are not highly sensitive to different prior settings. Their posterior distributions remain relatively stable across priors, showing strong data influence on these parameters.
- The posterior for sigma remains highly sensitive to the prior variance, suggesting that the data has weak influence on the variance of the posterior. This indicates that the residual variance is not well-determined by the data and 
relies heavily on prior information.


### 10.1.3. Multiple covariates for different kind of energy production:


```{r}
priors_lin_reg_multiple_fix <- list(
  i250_beta5_sigma10 = c(
    prior(normal(0, 250), class = "b", coef = "Intercept"),
    prior(normal(0, 5), class = "b", coef = "FossilFuelsPeatProductionShare"),
    prior(normal(0, 5), class = "b", coef = "NuclearEnergyProductionShare"),
    prior(normal(0, 5), class = "b", coef = "NetImportsProductionShare"),
    prior(normal(0, 5), class = "b", coef = "OthersProductionShare"),
    prior(normal(0, 10), class = "sigma")
  ),
  i50_beta1_sigma1 = c(
    prior(normal(0, 50), class = "b", coef = "Intercept"),
    prior(normal(0, 1), class = "b", coef = "FossilFuelsPeatProductionShare"),
    prior(normal(0, 1), class = "b", coef = "NuclearEnergyProductionShare"),
    prior(normal(0, 1), class = "b", coef = "NetImportsProductionShare"),
    prior(normal(0, 1), class = "b", coef = "OthersProductionShare"),
    prior(normal(0, 1), class = "sigma")
  ),
  i250_beta10_sigma10 = c(
    prior(normal(0, 250), class = "b", coef = "Intercept"),
    prior(normal(0, 10), class = "b", coef = "FossilFuelsPeatProductionShare"),
    prior(normal(0, 10), class = "b", coef = "NuclearEnergyProductionShare"),
    prior(normal(0, 10), class = "b", coef = "NetImportsProductionShare"),
    prior(normal(0, 10), class = "b", coef = "OthersProductionShare"),
    prior(normal(0, 10), class = "sigma")
  )
)

prior_check(priors_lin_reg_multiple_fix, lin_reg_multiple_fix_fit)
```

From the density plots of the posterior distributions under different priors, 
we can see that all variables are still somewhat sensitive to the very informative
priors, but the changes in posterior distributions are not significant. However,
slope of `OthersProductionShare` and `sigma` are still very sensitive and dependent
on the priors.

## 10.2. Time series models {#ts-sensitivity}

On the plots below, you can see posteriors of each parameter for different prior configurations. For the prior configurations, we tried different standard deviations of the normal priors for each parameter. The AR regression coefficient was tried with stds of 0.5, 1, 2. The general mean intercept had stds of 100, 500, 1000. The covariate coefficients had stds of 5, 10, 100. And the variance prior had the stds of 10, 100, 200.

```{r}
priors_ar <- list(
  ar1_i1000_sigma10=c(
    prior(normal(0, 1), class = "ar"),
    prior(normal(0,1000), class="Intercept"),
    prior(normal(0, 10), class = "sigma")
  ),
  ar2_i500_sigma100=c(
    prior(normal(0, 2), class = "ar"),
    prior(normal(0,500), class="Intercept"),
    prior(normal(0, 100), class = "sigma")
  ),
  ar0.5_i100_sigma200=c(
    prior(normal(0, 0.5), class = "ar"),
    prior(normal(0, 100), class="Intercept"),
    prior(normal(0, 200), class = "sigma")
  )
)
prior_check(priors_ar, ar_fit) + labs(title="AR(1)")
```

As you can see, the posteriors of all paramters of AR(1) model are fairly robust to changes in prior variance, except for sigma which is clearly perturbed by the stronger priors.

```{r}
priors_ar_renewable <- list(
  ar1_i1000_b5_sigma_10=c(
  prior(normal(0, 1000), class = "b", coef = "Intercept"),
  prior(normal(0, 5), class = "b", coef = "RenewableProductionShare"),
  prior(normal(0, 10), class = "sigma"),
  prior(normal(0, 1), class = "ar")
  ),
  ar2_i500_b10_sigma_100=c(
  prior(normal(0, 500), class = "b", coef = "Intercept"),
  prior(normal(0, 10), class = "b", coef = "RenewableProductionShare"),
  prior(normal(0, 100), class = "sigma"),
  prior(normal(0, 2), class = "ar")
  ),
  ar0.5_i100_b100_sigma_200=c(
  prior(normal(0, 100), class = "b", coef = "Intercept"),
  prior(normal(0, 100), class = "b", coef = "RenewableProductionShare"),
  prior(normal(0, 200), class = "sigma"),
  prior(normal(0, 0.5), class = "ar")
  )
)
prior_check(priors_ar_renewable, ar_renewable_fit) + labs(title="AR(1) + Renewable")
```
For the model with AR(1) component and renewable energy share covariates we see that all of the parameters' posteriors are sensitive to the prior variance to some extent which indicates that the amount of data is not sufficient to overcome strongly informative priors.

```{r}
priors_ar_fossil <- list(
  ar1_i1000_b5_sigma_10=c(
  prior(normal(0, 1000), class = "b", coef = "Intercept"),
  prior(normal(0, 5), class = "b", coef = "FossilFuelsPeatProductionShare"),
  prior(normal(0, 10), class = "sigma"),
  prior(normal(0, 1), class = "ar")
  ),
  ar2_i500_b10_sigma_100=c(
  prior(normal(0, 500), class = "b", coef = "Intercept"),
  prior(normal(0, 10), class = "b", coef = "FossilFuelsPeatProductionShare"),
  prior(normal(0, 100), class = "sigma"),
  prior(normal(0, 2), class = "ar")
  ),
  ar0.5_i100_b100_sigma_200=c(
  prior(normal(0, 100), class = "b", coef = "Intercept"),
  prior(normal(0, 100), class = "b", coef = "FossilFuelsPeatProductionShare"),
  prior(normal(0, 200), class = "sigma"),
  prior(normal(0, 0.5), class = "ar")
  )
)
prior_check(priors_ar_fossil, ar_fossil_fuel_fit) + labs(title="AR(1) + Fossil")
```


# 11. Model comparison (e.g. with LOO-CV)

Finally, let's compare all our models using leave-future-out cross validation (LFO). We reserve 10 data points for the models to learn and leave the next points out, adding them one by one. We perform LFO using Pareto sampling to avoid fitting the model several times, for that we create a pareto_lfo function.

```{r}
log_sum_exp <- function(x) {
  max_x <- max(x)  
  max_x + log(sum(exp(x - max_x)))
}

# more stable than log(mean(exp(x)))
log_mean_exp <- function(x) {
  log_sum_exp(x) - log(length(x))
}

# compute log of raw importance ratios
# sums over observations *not* over posterior samples
sum_log_ratios <- function(loglik, ids = NULL) {
  if (!is.null(ids)) loglik <- loglik[, ids, drop = FALSE]
  rowSums(loglik)
}

pareto_lfo <- function(fit, L, data, name) {
  M <- 1
  df <- data
  plot_ks <- function(ks, ids, thres = 0.6) {
  dat_ks <- data.frame(ks = ks, ids = ids)
  ggplot(dat_ks, aes(x = ids, y = ks)) + 
    geom_point(aes(color = ks > thres), shape = 3, show.legend = FALSE) + 
    geom_hline(yintercept = thres, linetype = 2, color = "red2") + 
    scale_color_manual(values = c("cornflowerblue", "darkblue")) + 
    labs(x = "Data point", y = "Pareto k") + 
    ylim(-0.5, 1.5)
}
  
  N <- length(data$CO2EquivEmission)
  k_thres <- 0.7
  approx_elpds_1sap <- rep(NA, N)

# initialize the process for i = L
past <- 1:L
oos <- L + 1
df_past <- df[past,]
df_oos <- df[c(past, oos),]
fit_past <- update(fit, newdata = df_past, recompile = FALSE, refresh=0)
loglik <- log_lik(fit_past, newdata = df_oos, oos = oos)
approx_elpds_1sap[L + 1] <- log_mean_exp(loglik[, oos])

# iterate over i > L
i_refit <- L
refits <- L
ks <- NULL
for (i in (L + 1):(N - 1)) {
  past <- 1:i
  oos <- i + 1
  df_past <- df[past,]
  df_oos <- df[c(past, oos),]
  loglik <- log_lik(fit_past, newdata = df_oos, oos = oos)
  
  logratio <- sum_log_ratios(loglik, (i_refit + 1):i)
  psis_obj <- suppressWarnings(psis(logratio))
  k <- suppressWarnings(pareto_k_values(psis_obj))
  ks <- c(ks, k)
  if (k > k_thres) {
    # refit the model based on the first i observations
    i_refit <- i
    refits <- c(refits, i)
    fit_past <- update(fit_past, newdata = df_past, recompile = FALSE, refresh=0)
    loglik <- log_lik(fit_past, newdata = df_oos, oos = oos)
    approx_elpds_1sap[i + 1] <- log_mean_exp(loglik[, oos])
  } else {
    lw <- weights(psis_obj, normalize = TRUE)[, 1]
    approx_elpds_1sap[i + 1] <- log_sum_exp(lw + loglik[, oos])
  }
} 
    approx_elpd <- sum(approx_elpds_1sap, na.rm = TRUE)
    cat(
    name, " ", "approx LFO ", approx_elpd, "\n"
    )
    cat("Using threshold ", k_thres, 
      ", model was refit ", length(refits), 
      " times, at observations", refits, "\n")
}
```

## 11.1 Linear models
```{r}
pareto_lfo(lin_reg_renewable_fit, 10, emission, "LinReg Renewable")
```
```{r}
pareto_lfo(lin_reg_fossil_fuel_fit, 10, emission, "LinReg FossilFuels")
```
```{r}
pareto_lfo(lin_reg_multiple_fix_fit, 10, emission, "LinReg Multiple fix")
```

## 11.2 Autoregressive models
```{r}
pareto_lfo(ar_fit, 10, emission, "AR(1)")
```
```{r}
pareto_lfo(ar_renewable_fit, 10, emission, "AR(1) + Renewable")
```
```{r}
pareto_lfo(ar_fossil_fuel_fit, 10, emission, "AR(1) + Fossil")
```


# 12. Discussion of Issues and Potential Improvements

## 12.1. Issues

The first issue is that our dataset is small and lacks variability. This restricts 
us to using relatively basic models and prevents the application of more advanced 
techniques, such as hierarchical models.

In addition, the available time series is fairly short, making it challenging to 
accurately assess the model's predictive performance. Due to the short time series, 
our analysis was limited to understanding emissions through their time-series 
characteristics and their dynamics with production shares of different energy sources. 
We were unable to make meaningful predictions or forecasts about future emissions
as we stated earlier in the introduction.

## 12.2. Potential Improvements

First of all, future work could focus more about explaining the bimodal replicates
distribution of linear models using fossil fuels production share. One potential
explanation is the presence of non-linearity that our linear models could not explain.
Therefore, one can try to use non-linear models or incorporate non-linear terms 
into the existing linear models. This approach could capture more complex 
relationships between emissions and production shares of energy sources,
especially since the dynamics might not be strictly linear.

Secondly, one can expanding the scope of the analysis to include predictors 
beyond the production share of energy sources. For example, consumption data, 
especially the share of renewable energy in consumption, could provide additional 
insights. Other potential predictors might include economic variables, 
government policies, or technological changes influencing energy production 
and consumption, which are harder to give concrete examples, but intuitively
they should be able to contribute to the predictive performance.

Lastly, being able to extend the dataset will open many more doors. Acquiring 
more data, both in terms of time (longer time series) and scope (broader 
variability), would allow for the use of advanced models such as hierarchical
models and improve the robustness of the analysis.

# 13. Conclusion what was learned from the data analysis

From the data analysis, we learned the following key insights:
- The share of fossil fuels and peat in production is a very strong predictor for the emission coefficient. Across all models, this covariate consistently demonstrated its dominant influence on emissions, with robust and stable posterior estimates.
- Contrary to our initial expectations, the share of renewable energy in production is not a strong predictive covariate for emissions. Models that included renewable production share as a primary predictor did not perform well, showing weaker alignment with the observed data.

The analysis underscores that the share of fossil fuels remains the dominant driver of CO2-equivalent emissions. Renewable production share, though an important policy focus, does not currently hold strong predictive power for emissions in the dataset analyzed. 

# 14. Self-reflection of what the group learned while making the project

This project was our first experience using `brms` and `Stan` with real-world data, and we’re very happy with how much we learned throughout the process. One of the major challenges we faced was dealing with MCMC divergence for the first time. Debugging and fixing it gave us valuable insight into how our assumptions and model formulations impact the results. It was rewarding to revisit and refine our initial model to resolve the issue.

Working with real-world data also highlighted how different it is compared to the clean and structured datasets we typically encounter in homework. Our posterior predictive checks weren’t as good as we’re used to seeing, and the dataset we had was limited in both size and scope. We now recognize the importance of having access to richer data and additional predictors, as this could significantly improve the analysis.

That said, we’re proud of the model we built using the techniques learned in this course. Despite the challenges, we were able to identify key insights, such as the dominant role of fossil fuel production in predicting emissions. We're also grateful for existence of such modern software like `brms`, which has removed all the heavy-duty coding work width such complex samplers and algorithms, and allowed us to focus on interpreting the results and understanding the data.







