---
title: "CS-E5710 Bayesian Data Analysis - PROJECT NAME HERE" # TODO: project name here
author: "Egor Eremin & Tam Nguyen"
output: 
  pdf_document:
    toc: yes
    toc_depth: 1
urlcolor: blue
editor_options: 
  markdown: 
    wrap: 72
---

```{r include=FALSE, results='hide'}
# Install/import packages
if (!require(readr)) {
    install.packages("readr")
    library(readr)
}

if (!require(tidybayes)) {
    install.packages("tidybayes")
    library(tidybayes)
}

if (!require(rlang)) {
    install.packages("rlang")
    library(rlang)
}

if (!require(purrr)) {
    install.packages("purrr")
    library(purrr)
}

if (!require(brms)) {
    install.packages("brms")
    library(brms)
}

if(!require(ggplot2)){
    install.packages("ggplot2")
    library(ggplot2)
}

if(!require(ggplot2)){
    install.packages("ggplot2")
    library(ggplot2)
}

if(!require(bayesplot)){
    install.packages("bayesplot")
    library(bayesplot)
}

if(!require(ggpubr)){
    install.packages("ggpubr")
    library(ggpubr)
}

if (!require(loo)) {
  install.packages("loo")
  library(loo)
}

if (!require(priorsense)) {
  install.packages("priorsense")
  library(priorsense)
}

if (!require(reshape2)) {
  install.packages("reshape2")
  library(reshape2)
}

if(!require(cmdstanr)){
    install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
    library(cmdstanr)
}
cmdstan_installed <- function(){
  res <- try(out <- cmdstanr::cmdstan_path(), silent = TRUE)
  !inherits(res, "try-error")
}
if(!cmdstan_installed()){
    install_cmdstan()
}

```

```{r include=FALSE, results='hide'}
# General settings (e.g., seed, package options setting)
seed <- 42

set.seed(seed)
```


```{r}
emission <- read_csv("../data_processing/data/data.csv", show_col_types = FALSE)
head(emission)
```

# 3 & 4. Model description & priors: Autoregressive models

In addition to the linear models, we will also try using models with an AR component on the CO2 emission coefficient to capture possible "inertia" of CO2 emissions caused by changes in the energy infrastructure, as well as other factors besides the change in the distribution of the energy production methods such as improvements in energy production technology reducing emissions in general.

The plain autoregressive model is given by
$$
y_i = \alpha + \sum_{j=1}^p \phi_j y_{i-j} + \varepsilon 
$$
where:
- $y_i$ is the $i$th observation of the target variable $CO2EquivEmission$,
- $\alpha \in \mathbb{R}$ is the unknown scalar intercept,
- $\phi_j \in \mathbb{R}$ are the unknown regression coefficients for each lag,
- $\varepsilon \sim \mathcal{N}(0, \sigma^2)$ the noise, and
- $p$ is the order of the autoregressive model.

The Bayesian formulation of the model is practically the same as that for the linear models, having added AR(1) component to the mean of the model distribution. 

We are going to add only an AR component of order 1, i.e. we use the previous year's observation to predict the next one. The model's mean is then given by:

$$
\text{CO2EquivEmission} = \alpha + AR(1)
$$
Using one covariate $RenewableProductionShare$ and an AR(1):

$$
\text{CO2EquivEmission} = \alpha + \beta \cdot \text{RenewableProductionShare} + AR(1)
$$

Using one covariate $FossilFuelsPeatProductionShare$ and an AR(1):

$$
\text{CO2EquivEmission} = \alpha + \beta \cdot \text{FossilFuelsPeatProductionShare} + AR(1)
$$

We use the same prior choices as for the linear models. The AR regression coefficient has the following weakly informative prior:
$$
\phi \sim \mathcal{N}(0, 1)
$$


# 5. brms/Stan code & # 6. MCMC inference

We have run the MCMC inference using the same options as for the linear models. We performed 2000 iterations with 1000 left out for warmup and 4 MCMC chains. 

## 5.1 AR(1):
```{r results='hide'}
ar_formula <- bf(
  CO2EquivEmission ~ ar(time = Year, p=1), family="Gaussian"
)
get_prior(ar_formula, data=emission)
ar_priors <- c(
  prior(normal(0, 1), class = "ar"),
  prior(normal(0, 500), class="Intercept"),
  prior(normal(0, 10), class = "sigma")
)

ar_fit <- brm(
  formula = ar_formula, 
  prior = ar_priors, 
  data=emission, 
  iter = 2000,
  warmup = 1000,
  chains = 4,
  seed = seed
  )
```

## 5.2. One covariate $RenewableProductionShare$ and AR(1):

```{r results='hide'}
ar_renewable_formula <- bf(
  CO2EquivEmission ~ RenewableProductionShare + ar(time = Year, p=1),
  family = "gaussian",
  center = FALSE
)

ar_renewable_priors <- c(
  prior(normal(0, 500), class = "b", coef = "Intercept"),
  prior(normal(0, 10), class = "b", coef = "RenewableProductionShare"),
  prior(normal(0, 10), class = "sigma"),
  prior(normal(0, 1), class = "ar")
  )

ar_renewable_fit <- brm(
  formula = ar_renewable_formula,
  prior = ar_renewable_priors,
  data = emission,
  iter = 2000,
  warmup = 1000,
  chains = 4,
  seed = seed
)
```

## 5.3. One covariate $FossilFuelsPeatProductionShare$ and AR(1): 

```{r results='hide'}
ar_fossil_fuel_formula <- bf(
  CO2EquivEmission ~ FossilFuelsPeatProductionShare + ar(time = Year, p=1),
  family = "gaussian",
  center = FALSE
)

ar_fossil_fuel_priors <- c(
  prior(normal(0, 500), class = "b", coef = "Intercept"),
  prior(normal(0, 10), class = "b", coef = "FossilFuelsPeatProductionShare"),
  prior(normal(0, 10), class = "sigma"),
  prior(normal(0, 1), class = "ar")
)

ar_fossil_fuel_fit <- brm(
  formula = ar_fossil_fuel_formula,
  prior = ar_fossil_fuel_priors,
  data = emission,
  iter = 2000,
  warmup = 1000,
  chains = 4,
  seed = seed,
  control=list(adapt_delta = 0.99)
)
```


# 7. Convergence diagnosis

MCMC showed no signs of divergence: we had no divergent transitions, the $\hat R$ values for all of our models are <1.05. Even though the effective samples sizes differ from the number of samples significantly, they still show a decent amount indicating that the sampling process had no serious divergence issues and can be used for analysis.

```{r}
summary(ar_fit)

summary(ar_renewable_fit)

summary(ar_fossil_fuel_fit)
```


# 8. Posterior predictive checks

We can see that the plain AR model, expectedly, performed the worst: Some draws indicate bimodal distribtuions, while other draws have too light tails and overestimate density for the middle values. 

AR model with a renewable fuel share covariate performs significantly better as its posterior matches that of data more closely. We still have a problem with tails, where the left side of the posterior is too heavy-tailed and the right side still undersetimates thickness for some draws. We also still observe overestimation of the middle values density in the posterior. 

AR model with the fossil fuel covariate shows the best fit of the posterior to data. The model still produces a bimodal distribution sometimes, as well as still underestimates the right tail values. 

```{r}
plot1 <- pp_check(ar_fit) + labs(title="AR(1)")

plot2 <- pp_check(ar_renewable_fit)+ labs(title="AR(1) + Renewable")

plot3 <- pp_check(ar_fossil_fuel_fit)+ labs(title="AR(1) + Fossil")

ggarrange(plot1, plot2, plot3, common.legend=TRUE, ncol=3) 
```


# 9. Predictive performance assessment

Below you can find the plots for models' predicitons. The light blue line is the mean, dark-blue shaded area shows the 50% CI and the light-blue area shows the 95

```{r}
p1 <- pp_check(ar_fit, type = "ribbon") + labs(title="AR(1)")

p2 <- pp_check(ar_renewable_fit, type = "ribbon") + labs(title="AR(1) + Renewable")

p3 <- pp_check(ar_fossil_fuel_fit, type = "ribbon") + labs(title="AR(1) + Fossil")

ggarrange(p1, p2, p3, common.legend=TRUE, ncol=3)
```

# 10. Sensitivity analysis

```{r}
prior_check <- function(priors, model) {
  draws <- priors %>%
  map(~ as.matrix(update(ar_fit, prior=.x, seed=seed)))
  
  draws_to_plot <- function(draws) {
    params <- head(colnames(draws[[1]]), -2)
    data <- list()
    for (p in params) {
        for (name in names(draws)) {
          t <- draws[[name]][,p]
          for (s in 1:length(t)) {
            data[[p]][[name]][[s]] = t[[s]]
          }
        }
    }
    data
  }
  dt <- draws_to_plot(draws)
  
  plot_for_loop <- function(df, x_var) {
    df.long <- melt(df)
      ggplot(df.long, aes(x= value, fill=L1)) +
      geom_density(alpha=0.7) +
      labs(x = x_var) +
      theme_classic(base_size = 12)
  }
  
  plot_list <- list()
  params <- head(colnames(draws[[1]]), -2)
  for (p in params) {
    plot_list[[length(plot_list)+1]] <- plot_for_loop(dt[[p]], p)
  }
  
  # Combine all plots
  ggarrange(plotlist=plot_list, common.legend=TRUE)
}
```
```{r}
priors_ar <- list(
  ar1_i1000_sigma10=c(
    prior(normal(0, 1), class = "ar"),
    prior(normal(0,1000), class="Intercept"),
    prior(normal(0, 10), class = "sigma")
  ),
  ar2_i500_sigma100=c(
    prior(normal(0, 2), class = "ar"),
    prior(normal(0,500), class="Intercept"),
    prior(normal(0, 100), class = "sigma")
  ),
  ar3_i250_sigma200=c(
    prior(normal(0, 3), class = "ar"),
    prior(normal(0, 250), class="Intercept"),
    prior(normal(0, 200), class = "sigma")
  )
)

priors_ar_renewable <- list(
  ar1_i1000_b50_sigma_10=c(
  prior(normal(0, 1000), class = "b", coef = "Intercept"),
  prior(normal(0, 50), class = "b", coef = "RenewableProductionShare"),
  prior(normal(0, 10), class = "sigma"),
  prior(normal(0, 1), class = "ar")
  ),
  ar2_i500_b100_sigma_100=c(
  prior(normal(0, 500), class = "b", coef = "Intercept"),
  prior(normal(0, 100), class = "b", coef = "RenewableProductionShare"),
  prior(normal(0, 100), class = "sigma"),
  prior(normal(0, 2), class = "ar")
  ),
  ar3_i250_b200_sigma_250=c(
  prior(normal(0, 250), class = "b", coef = "Intercept"),
  prior(normal(0, 200), class = "b", coef = "RenewableProductionShare"),
  prior(normal(0, 250), class = "sigma"),
  prior(normal(0, 3), class = "ar")
  )
)

priors_ar_fossil <- list(
  ar1_i1000_b50_sigma_10=c(
  prior(normal(0, 1000), class = "b", coef = "Intercept"),
  prior(normal(0, 50), class = "b", coef = "FossilFuelsPeatProductionShare"),
  prior(normal(0, 10), class = "sigma"),
  prior(normal(0, 1), class = "ar")
  ),
  ar2_i500_b100_sigma_100=c(
  prior(normal(0, 500), class = "b", coef = "Intercept"),
  prior(normal(0, 100), class = "b", coef = "FossilFuelsPeatProductionShare"),
  prior(normal(0, 100), class = "sigma"),
  prior(normal(0, 2), class = "ar")
  ),
  ar3_i250_b200_sigma_250=c(
  prior(normal(0, 250), class = "b", coef = "Intercept"),
  prior(normal(0, 200), class = "b", coef = "FossilFuelsPeatProductionShare"),
  prior(normal(0, 250), class = "sigma"),
  prior(normal(0, 3), class = "ar")
  )
)
prior_check(priors_ar, ar_fit)
```
```{r}
prior_check(priors_ar_renewable, ar_renewable_fit)
```
```{r}
prior_check(priors_ar_fossil, ar_fossil_fuel_fit)
```

# 11. Model comparison (e.g. with LOO-CV)
```{r}
log_sum_exp <- function(x) {
  max_x <- max(x)  
  max_x + log(sum(exp(x - max_x)))
}

# more stable than log(mean(exp(x)))
log_mean_exp <- function(x) {
  log_sum_exp(x) - log(length(x))
}

# compute log of raw importance ratios
# sums over observations *not* over posterior samples
sum_log_ratios <- function(loglik, ids = NULL) {
  if (!is.null(ids)) loglik <- loglik[, ids, drop = FALSE]
  rowSums(loglik)
}

pareto_lfo <- function(fit, L, data) {
  M <- 1
  df <- data
  plot_ks <- function(ks, ids, thres = 0.6) {
  dat_ks <- data.frame(ks = ks, ids = ids)
  ggplot(dat_ks, aes(x = ids, y = ks)) + 
    geom_point(aes(color = ks > thres), shape = 3, show.legend = FALSE) + 
    geom_hline(yintercept = thres, linetype = 2, color = "red2") + 
    scale_color_manual(values = c("cornflowerblue", "darkblue")) + 
    labs(x = "Data point", y = "Pareto k") + 
    ylim(-0.5, 1.5)
}
  
  N <- length(data$CO2EquivEmission)
  loo_cv <- loo(log_lik(fit)[, (L + 1):N])
  k_thres <- 0.7
  approx_elpds_1sap <- rep(NA, N)

# initialize the process for i = L
past <- 1:L
oos <- L + 1
df_past <- df[past,]
df_oos <- df[c(past, oos),]
fit_past <- update(fit, newdata = df_past, recompile = FALSE, refresh=0)
loglik <- log_lik(fit_past, newdata = df_oos, oos = oos)
approx_elpds_1sap[L + 1] <- log_mean_exp(loglik[, oos])

# iterate over i > L
i_refit <- L
refits <- L
ks <- NULL
for (i in (L + 1):(N - 1)) {
  past <- 1:i
  oos <- i + 1
  df_past <- df[past,]
  df_oos <- df[c(past, oos),]
  loglik <- log_lik(fit_past, newdata = df_oos, oos = oos)
  
  logratio <- sum_log_ratios(loglik, (i_refit + 1):i)
  psis_obj <- suppressWarnings(psis(logratio))
  k <- pareto_k_values(psis_obj)
  ks <- c(ks, k)
  if (k > k_thres) {
    # refit the model based on the first i observations
    i_refit <- i
    refits <- c(refits, i)
    fit_past <- update(fit_past, newdata = df_past, recompile = FALSE, refresh=0)
    loglik <- log_lik(fit_past, newdata = df_oos, oos = oos)
    approx_elpds_1sap[i + 1] <- log_mean_exp(loglik[, oos])
  } else {
    lw <- weights(psis_obj, normalize = TRUE)[, 1]
    approx_elpds_1sap[i + 1] <- log_sum_exp(lw + loglik[, oos])
  }
} 
    approx_elpd <- sum(approx_elpds_1sap, na.rm = TRUE)
    cat(
    "approx LFO ", approx_elpd, "\n"
    )
    cat("Using threshold ", k_thres, 
      ", model was refit ", length(refits), 
      " times, at observations", refits)
    plot_ks(ks, (L + 1):(N - M))
}
```
```{r}
pareto_lfo(ar_fit, 5, emission)
pareto_lfo(ar_renewable_fit, 5, emission)
pareto_lfo(ar_fossil_fuel_fit, 5, emission)
```
